---
title: 'Acoustic pre-processing'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acoustic pre-processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following set of functions help to pre-process and organize audio and corresponding metadata. \code(`wt_audio_scanner`) scans a directory of audio files and prepares them in a tibble with WildTrax formatted columns. \code(`wt_run_ap`) allows you to generate acoustic indices and false-colour spectrograms from a \code(`wt_audio_scanner`) tibble, while \code{`wt_glean_ap`} wrangles the output into summary plots and LDFCs. \code(`wt_signal_level`) detects signals in audio based on amplitude thresholds. In conjunction, these tools allow you to select recordings parameterized to a specific study design. `vignette("linking-media-to-wildtrax")` will allow you to dive deeper in how to link the data to the WildTrax platform.

```{r, include = FALSE, echo = F, include = T, warning = F, message = F}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)

library(wildRtrax)
library(tidyverse)

```

## Scanning audio files from a directory

The `wt_audio_scanner()` function reads in audio files (either wac, wav or flac format) from a local directory and outputs useful metadata.

```{r, echo=TRUE, include=TRUE, eval=T, warning = F, message = F}
# Scan data

if (dir.exists("../")) {
  wt_audio_scanner(path = "../", file_type = "wav", extra_cols = T)
} else {
  'Can\'\t find this directory'
}

```

You might want to select recordings between certain times of day or year. 

```{r, echo=TRUE, include=F, eval=T, warning = F, message = F}
load("../book/book.RData")

```

```{r, echo=TRUE, include=TRUE, eval=T, warning = F, message = F}
files

```

Or filter recordings based on some criteria.

```{r, echo=TRUE, include=TRUE, eval=T, warning = F, message = F}
library(lubridate)

files %>%
  mutate(hour = hour(recording_date_time)) %>%
  filter(julian == 176, 
         hour %in% c(4:8))

```

## Running the QUT Ecoacoustics AnalysisPrograms software on a wt_* standard data set

The `wt_run_ap()` function allows you to run the QUT Analysis Programs [(AP.exe)](https://ap.qut.ecoacoustics.info/) on your audio data. AP generates acoustic index values and false-colour spectrograms for each audio minute of data. Note that you must have the AP program installed on your computer. See more here [(Towsey et al., 2018)](https://researchoutput.csu.edu.au/ws/portalfiles/portal/28556441/28544328_Published_article.pdf). 

```{r echo=TRUE, include=T, eval=F, warning = F, message = F}
#Use the wt_* tibble to execute the AP on the files

wt_run_ap(x = my_files, output_dir = paste0(root, 'ap_outputs'), path_to_ap = '/where/you/store/AP')

```

The use `wt_glean_ap` to plot the acoustic index and LDFC results

```{r, echo=T, include=T, eval=F, warning=F, message, prompt=T, comment=""}

wt_glean_ap()

```

## Applying a limited amplitude filter

We can use the `wt_signal_level()` function to search for sounds that exceed a certain amplitude threshold. 

```{r echo=TRUE, include=TRUE, eval=F}
# Example audio file
path_to_file <- b$file_path[3]

# Run
s <- wt_signal_level(path = path_to_file, 
                     fmin = 5000, 
                     fmax = 10000, 
                     threshold = 20, 
                     channel = 'left', 
                     aggregate = 5)

# Return a list object, with parameters stored
str(s)

# We can view the output:
s['output']
# We have eleven detections that exceeded this threshold.

```

## Linking files to WildTrax

Make tasks at any time using a `wt_*` standard data set.

```{r, include = T, eval=F, warning=F, message=F}
wt_make_aru_tasks(input = testff, task_method = "1SPT", task_length = 180)

```

The function `wt_songscope_tags` reformats the output obtained from a wildlife acoustics songscope recognizer. This transformation involves converting the recognizer tags into tags that do not have a method type. This makes it possible to upload each hit as a tag in a task.

```{r, include = T, eval=F, warning=F, message=F, prompt="", prompt=T}
wt_songscope_tags()

```

Similarly, the function `wt_kaleidoscope_tags` performs the same reformatting process, but with Kaleidoscope instead. It is worth noting that this function targeted for sonic and ultrasonic species upload.

```{r, include = T, eval=F, warning=F, message=F}
wt_kaleidoscope_tags()

```

