---
title: "Acoustic data wrangling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acoustic data wrangling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The recommended workflow to wrangle together data for analysis in `wildRtrax` is as follows. Once you have your data from `wt_download_report()`.

- Remove unneeded species: `wt_tidy_species()`
- Convert TMTT counts to numeric: `wt_replace_tmtt()`
- Convert the species to a wide format: `wt_make_wide()`
- Add additional [offsets](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12106) for the data `wt_qpad_offsets()`

```{r, echo = F, warning = F, message = F, eval = T}

library(wildRtrax)
library(tidyverse)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)

load("book.RData")

```

```{r, warning = F, message = F, eval = F, include = T, echo = T}
# Start by getting everything you need
Sys.setenv(WT_USERNAME = 'guest', WT_PASSWORD = 'Apple123')
wt_auth()
my_report <- wt_download_report(project_id = 379, sensor_id = 'ARU', reports = "main", weather_cols = F) %>%
  tibble::as_tibble()

head(my_report)

```

## Data wrangling

Now let's start cleaning things up a bit. We don't need certain abiotic codes so let's remove those with `wt_tidy_species()`,

```{r, eval = F, echo = T, warning = F, message = F}
my_tidy_data <- wt_tidy_species(my_report, remove = c("mammal", "amphibian", "abiotic", "insect", "unknown", zerofill=TRUE))

# Difference in rows
round((nrow(my_tidy_data)/nrow(my_report)),2)
```

convert TMTT counts to numeric,

```{r, eval = F, echo = T, warning = F, message = F}
my_tmtt_data <- wt_replace_tmtt(data = my_tidy_data, calc = "round")

```

and finally widen the data into a species matrix.

```{r, eval = F, echo = T, warning = F, message = F}
my_wide_data <- wt_make_wide(data = my_tmtt_data, sound = "all")

```

## Offsets

Now you can calculate statistical offsets to account for imperfection detection following the QPAD method (Solymos et al. 2013 in Methods in Ecology & Evolution).

```{r, eval = F, echo = T, warning = F, message = F}
my_offset_data <- wt_qpad_offsets(data = my_wide_data, species = "all", version = 3, together = TRUE)

```

## Occupancy modelling

You can also perform a single-season, single-species occupancy work flow using `wt_format_occupancy()` once the data is downloaded.

```{r, eval = F, echo = T, warning = F, message = F}
<<<<<<< HEAD
<<<<<<< HEAD
dat.occu <- wt_format_occupancy(my_report, species="CONI", siteCovs=NULL)
=======
dat.occu <- wt_format_occupancy(my_data, species="WTSP", siteCovs=NULL)
>>>>>>> 98cbec44583e541d85b7c568570b1ee7032c4a0f
=======
dat.occu <- wt_format_occupancy(my_report, species="CONI", siteCovs=NULL)
>>>>>>> 70c5a0272f45c44e5eb385b2c40ace4248e2f507
mod <- unmarked::occu(~ 1 ~ 1, dat.occu)
mod

```
