---
title: 'Acoustic data wrangling'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acoustic data wrangling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The recommended workflow to wrangle together data for analysis in `wildrtrax` is as follows. Once you have your data from `wt_download_report()`.

- Remove unneeded species: `wt_tidy_species()`
- Convert TMTT counts to numeric: `wt_replace_tmtt()`
- Convert the species to a wide format: `wt_make_wide()`

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)

load("acoustic_wrangling.RData")
#save.image("acoustic_wrangling.RData")
```

```{r setup, echo=FALSE, include=FALSE, eval=TRUE}
# Attach package
library(wildrtrax)
library(dplyr)

```

```{r, warning = F, message = F, eval = T, include = T, echo = T}
# Start by getting everything you need
Sys.setenv(WT_USERNAME = 'guest', WT_PASSWORD = 'Apple123')
wt_auth()
my_report <- wt_download_report(project_id = 620, sensor_id = 'ARU', reports = "main") 


```

## Data wrangling

Let's use some of the functions that are available to start cleaning up a data set. For this question, we don't need certain abiotic and mammal codes so let's remove those with `wt_tidy_species()`,

```{r, tidy-chunk, eval = T, echo = T, warning = F, message = F}
my_tidy_data <- wt_tidy_species(my_report, remove = c("mammal"), zerofill = F)

# Difference in rows
round((nrow(my_tidy_data)/nrow(my_report)),2)
```

So about 15% of detections were mammals. Next, convert TMTT (too many to tag) counts to numeric:

```{r, eval = T, echo = T, warning = F, message = F}
my_tmtt_data <- wt_replace_tmtt(data = my_tidy_data, calc = "round")

```

and finally, widen the data into a species matrix.

```{r, eval = T, echo = T, warning = F, message = F}
my_wide_data <- wt_make_wide(data = my_tmtt_data, sound = "all")

head(my_wide_data)

```

## Occupancy modelling

You can also perform a single-season, single-species occupancy work flow using `wt_format_occupancy()` once the data is downloaded.

```{r, eval = T, echo = T, warning = F, message = F}
dat.occu <- wt_format_occupancy(my_report, species="WTSP", siteCovs=NULL)
mod <- unmarked::occu(~ 1 ~ 1, dat.occu)
mod

```
