---
title: "Authenticating into WildTrax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{authenticating-into-wt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=TRUE, eval=FALSE}

library(wildRtrax)

```

## Set up credentials

To obtain an Auth0 token, you need to first store your WildTrax username and password as environment variables.

```{r echo=TRUE, eval=FALSE}
# Note that you need to use 'WT_USERNAME' and 'WT_PASSWORD'. 
# Sys.setenv(WT_USERNAME = "my_username", WT_PASSWORD = "my_password")

```

## Authenticate

Next, you use the `wt_auth()` function to authenticate.

```{r echo=TRUE, eval=FALSE}
# Authenticate
# wt_auth()

```

The Auth0 token you obtained will last for 16 hours. After that time, you will need to re-authenticate.

## Making API calls

Once authenticated, you can now use various functions that call upon the WildTrax API. For instance, you can use `wt_get_download_summary()` to see basic metadata about projects that you can download data for.  

```{r echo=TRUE, eval=FALSE}
# wt_get_download_summary(sensorId = "ARU")

```




