---
title: 'Bat data workflows'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bat data workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Bat data concepts

Autonomous recording units (ARUs) deployed for bat monitoring differ from those used for birds or amphibians in one key way: they record in the **ultrasonic** range. Most bat echolocation calls fall roughly between 9 kHz and 120 kHz, well above the upper limit of human hearing (~20 kHz). Because these calls are inaudible to people, specialized hardware and sample rates (typically 192–500 kHz) are required to capture them without aliasing, and specialized software is required to
interpret them afterward. This creates a distinct workflow compared to sonic (audible-range) ARU data:

- **Recording**: ultrasonic microphones and higher sample rates capture full-spectrum or zero-crossing audio of bat calls.
- **Classification**: because calls are short (often <100 ms) and highly variable between species, automated classifiers are used to scan recordings and assign a likely species identity to each detected call or file, rather than relying solely on human aural identification.
- **Verification**: classifier output is typically treated as a first-pass filter, with acoustic experts confirming or correcting species calls, especially for rare, at-risk, or acoustically similar species.
- **Standardization**: outputs from different classifiers use different file structures and species codes, so `wildrtrax` provides functions to translate these into a single, consistent tag format that can be synced with recordings already uploaded to [WildTrax](https://www.wildtrax.ca/).

## Bat classifiers

`wildrtrax` supports formatting output from three widely used bat classification tools:

| Classifier | Developer | Status | `wildrtrax` function |
|---|---|---|---|
| Sonobat | SonoBat | Actively maintained | *(see below)* |
| Kaleidoscope | Wildlife Acoustics | Actively maintained | `wt_kaleidoscope_tags()` |
| Songscope | Wildlife Acoustics | Deprecated | `wt_songscope_tags()` |

`wt_kaleidoscope_tags()` and `wt_songscope_tags()` take the classification output files produced by each program (typically a CSV or set of per-file classification results) and reformat them into the WildTrax tag schema; species code, confidence, start/end time, and frequency bounds per detection, so that tags line up correctly with their corresponding recordings once uploaded. A minimal workflow looks like:

```{r kaleidoscope-example}
library(wildrtrax)

# Point to the Kaleidoscope classifier output (e.g. a summary CSV)
kaleido_tags <- wt_kaleidoscope_tags(input = "path/to/kaleidoscope_output.csv")

# Inspect the standardized tag table before uploading
head(kaleido_tags)
```

```{r songscope-example}
# Songscope is deprecated by Wildlife Acoustics but historical
# datasets processed with it can still be brought into WildTrax
songscope_tags <- wt_songscope_tags(input = "path/to/songscope_output")
```

Because Songscope is no longer actively developed or supported by Wildlife Acoustics, `wt_songscope_tags()` exists primarily to support legacy data, new projects should generally classify with Kaleidoscope or Sonobat instead.

## GUANO metadata to tags

[GUANO](https://guano-md.org/) (**G**rand **U**nified **A**coustic **N**otation **O**ntology) is a metadata standard for bat acoustic recordings. Rather than living in a separate file, GUANO metadata is embedded directly in the WAV file itself, in a dedicated metadata chunk. This can include information such as:

- Species classification and classifier-assigned confidence
- Recording location (latitude/longitude), date, and time
- Equipment and firmware details
- Notes added manually by an analyst or automatically by classifier software (e.g. Kaleidoscope and some Sonobat workflows write GUANO tags back into the original recordings)

Because this metadata travels with the file rather than in a separate results table, it needs its own extraction step. The `wt_guano_tags()` function reads the embedded GUANO metadata from a set of recordings and exports it into the same standard WildTrax tag template used by the classifier functions above, ready for upload and synchronization with your recordings in WildTrax.

```{r guano-example}
guano_tags <- wt_guano_tags(input = "path/to/recordings/")

head(guano_tags)
```

## Putting it together: a typical bat workflow

1. **Deploy** ultrasonic ARUs and collect full-spectrum recordings.
2. **Classify** recordings using Kaleidoscope, Sonobat, or (for legacy data) Songscope.
3. **Standardize** the classifier output with the matching `wildrtrax` function (`wt_kaleidoscope_tags()`, `wt_songscope_tags()`), or extract embedded metadata directly with `wt_guano_tags()` if species information was written back into the files as GUANO.
4. **Upload** recordings to WildTrax, then **sync** the standardized tag table so that species detections line up with the correct recording and timestamp.
5. **Verify** classifier calls in WildTrax, correcting or confirming identifications as needed, particularly for species at risk or easily confused call types.

## A note on species identification confidence

Automated bat classifiers assign a confidence score (or probability) to each species call, and accuracy varies substantially by species, call quality, and background noise. It's good practice to treat classifier output as a starting point rather than a final identification, especially for species with overlapping call characteristics, and to have qualified personnel review calls flagged as at-risk or low-confidence before they are finalized in WildTrax.
