---
title: 'Camera data wrangling'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Camera data wrangling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=F, eval=F, message=TRUE, include=TRUE}

knitr::opts_chunk$set(echo = TRUE)
```

The use of camera traps in ecological studies has become increasingly popular for monitoring wildlife. Managing and analyzing camera trap data efficiently is important for extracting meaningful and accurate insights. In this vignette, we will explore how to perform camera data wrangling, specifically focusing on the ABMI's [Ecosystem Health 2014 data set](https://portal.wildtrax.ca/home/camera-deployments.html?sensorId=CAM&projectId=205).

Use the `wt_download_report()` function to retrieve the main report for the CAM sensor in the Ecosystem Health 2014 project:

```{r, echo=TRUE, eval=FALSE, message=TRUE, include=TRUE}

eh14_raw <- wt_download_report(
  project_id = 205, 
  sensor_id = "CAM",
  report = "main", 
  weather_cols = FALSE
)

```

Evaluate **independent detections** with `wt_ind_detect()`:

```{r echo=TRUE, eval=FALSE, message=TRUE, include=TRUE}

# Back to the Ecosystem Health 2014 data.

eh14_detections <- wt_ind_detect(
  x = eh14_raw, 
  threshold = 30,
  units = "minutes",
  remove_human = TRUE, 
  remove_domestic = TRUE 
)

```

```{r echo=TRUE, eval=F, message=TRUE, include=TRUE}

glimpse(eh14_detections, width = 75) 

```

**313** independent detections in this data set, when using a threshold of 30 minutes.

The output from `wt_ind_detect()` gave us some useful information. But we probably need to do additional wrangling for our data to be in the proper format for certain modeling techniques (e.g. habitat modeling, occupancy). For example, we want to evaluate the number of detections in a specified time interval (e.g. daily, weekly, or monthly), *including zeroes*. 

## Summarise your camera data

With `wt_summarise_cam()` you can get:

+ The output from `wt_ind_detect()` (e.g. the object `eh14_detections`)
+ Your raw data (e.g. the object `eh14_raw`)
+ The time interval you're interested in (e.g. weekly)
+ The variable you're interested in (e.g. detections, presence/absence)
+ The desired output format ('wide' or 'long')

```{r eval=FALSE, echo=TRUE, include=TRUE}
# A call to `wt_summarise_cam()`:

eh14_summarised <- wt_summarise_cam(
  # Supply your detection data
  detect_data = eh14_detections,
  # Supply your raw image data
  raw_data = eh14_raw,
  # Now specify the time interval you're interested in 
  time_interval = "week",
  # What variable are you interested in?
  variable = "detections",
  # Your desired output format (wide or long) 
  output_format = "wide" 
)

```

## The ultimate pipeline

```{r eval=FALSE, echo=TRUE, include=TRUE}

library(wildRtrax)

Sys.setenv(WT_USERNAME = "*****",
           WT_PASSWORD = "*****")
wt_auth()


projects <- wt_get_download_summary("CAM") %>%
  filter(project == "ABMI Ecosystem Health 2014") %>%
  select(project_id) %>%
  pull()

raw_data <- map_dfr(.x = projects, 
                    .f = ~wt_download_report(.x, "CAM", weather_cols = F, reports = "main")

summarised <- wt_ind_detect(raw_data, 30, "minutes") %>%
  wt_summarise_cam(raw_data, "day", "detections", "long")

```

And now you can get straight into the science!
