---
title: "Concepts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{concepts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE, include=FALSE, eval=TRUE}
library(wildRtrax)
```

## Introduction

### What is `wildRtrax`?

`wildRtrax`, pronounced *'wilder tracks'*, is an R package for ecologists and advanced users who work with environmental sensors such as autonomous recordings units (ARUs) and remote cameras. It contains functions designed to meet most needs in order to organize, analyze and standardize data to the [WildTrax](https://wwww.wildtrax.ca) infrastructure. `wildRtrax` is self-contained and must be run under an R statistical environment, and it also depends on many other R packages. `wildRtrax` is free software and distributed under [MIT License (c) 2020](https://github.com/ABbiodiversity/wildRtrax/blob/master/LICENSE).

### What is **WildTrax**?

[**WildTrax**](https://www.wildtrax.ca) is a web-enabled portal designed to manage, store, process, share and discover environmental sensor data and the  **biological data** extracted from the media. WildTrax was developed by the [Alberta Biodiversity Monitoring Institute](https://abmi.ca) and the [Bioacoustic Unit](https://bioacoustic.abmi.ca). `wildRtrax` serves as a parallel design and indicator to analytics and functions that can be developed in **WildTrax**. 

### What R packages does `wildRtrax` depend on?

`wildRtrax` depends on a multitude of packages to provide flexible routines and work flows for data management. [`tidyverse`](https://tidyverse.org) for piping functions, standard grammar and tidy data manipulation, [`furrr`](https://davisvaughan.github.io/furrr/) and [`doParallel`](https://cran.r-project.org/web/packages/doParallel/index.html) for parallel computing, and acoustic analysis packages: [`bioacoustics`](https://cran.r-project.org/web/packages/bioacoustics/index.html), [`tuneR`](https://cran.r-project.org/web/packages/tuneR/index.html), [`seewave`](https://cran.r-project.org/web/packages/seewave/seewave.pdf). Certain functions are indebted to the [QUT Ecoacoustics Analysis Software](https://github.com/QutEcoacoustics/audio-analysis) software package as well. 

### How do I report a bug in `wildRtrax`?

If you think you have found a bug in `wildRtrax`, you should report it to developers or maintainers. Please do not send bug reports to R mailing lists, since `wildRtrax` is not a standard R package. The preferred forum to report bugs is [GitHub](https://github.com/ABbiodiversity/wildRtrax/issues).  Here is what is required in order to report a bug - issues are welcome and encouraged and are the only way to make `wildRtrax` non-buggy:

* The bug report should be so detailed that the bug can be replicated and corrected
* Send an example that causes a bug
* Send a minimal data set as well if it is not available in R
* Paste the output or error message in your message
* Specify which version of `wildRtrax` you used

### Can I contribute to `wildRtrax`?

Yes! `wildRtrax` is dependent on user contribution and all feedback is welcome. If you have problems with `wildRtrax`, it may be as simple as incomplete documentation. Feature requests also are welcome, but they are not necessarily fulfilled. A new feature will be added if it is easy to do and it looks useful to the user base of the package, or if you submit fully annotated code.

See [here](https://abbiodiversity.github.io/wildRtrax/CONTRIBUTING.html) for more information. 

### Disclaimers

The pronoun “you” throughout these articles refers to the reader. “We” refers to the `wildRtrax` team in general.
