---
title: "linking-media-to-wildtrax"
author: "Alex MacPhail"
date: "09/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Linking files and creating templates for upload to [WildTrax](https://www.wildtrax.ca)

There are three phases of getting data into WildTrax once you have your [organization](https://www.wildtrax.ca/home/wildtrax-guide/2-Organizations.html) and [project](https://www.wildtrax.ca/home/wildtrax-guide/3-Projects.html) setup: uploading recordings, uploading tasks and uploading tags (see more in [ARU projects](https://www.wildtrax.ca/home/wildtrax-guide/3-Projects/3.2-ARU-projects.html) in WildTrax: The Definitive Guide).

If you've used `wt_audio_scanner` to scan through a series of nested directories for audio files, link them together.

```{r, eval = F, echo = T, include = T, warning = F, message = F}
#An example for a unix scneario

# Root directory
root <- '/volumes/GoogleDrive/Shared drives/wildRtrax/data/example'

# Scan data
my_files <- wildRtrax::wt_audio_scanner(path = root, file_type = "all", safe_scan = T, extra_cols = T, tz = "US/Mountain")

# View
head(my_files)

```

Now you should have all the links to the files across all directories in one folder. This is useful if you use a nested directory structure for your files (e.g. folders organized by location or visit).

The next step is to create the tasks and format the columns to the upload template. 


```{r}
#Create the template
#how to make tasks

```

