---
title: "Tutorials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{vembedr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

This vignette provides a series of additional tutorials to help you with using wildRtrax. 

# Webinars 

## wildRtrax 1.1.0 Public Webinar

In this webinar, Alex MacPhail, Marcus Becker and Elly Knight, take you on a tour of the main components of the package, including authenticating into [WildTrax](https://www.wildtrax.ca) directly in R, downloading data reports as data frames, and an introduction to various functions that aid in downstream analyses of both ARU and remote camera data.

```{r, echo=F}
vembedr::embed_url("https://www.youtube.com/watch?v=bckccZFRq7s&ab_channel=-abmicoms")

```

# Acoustic data

## Detect missing species using BirdNET

[BirdNET](https://birdnet.cornell.edu/) is an avian multi-species classifier. WildTrax utilizes the [BirdNET API](https://github.com/kahst/BirdNET-Analyzer) to automatically detect species in recordings uploaded to projects once the task is Transcribed. The results can then be retrieved using `wt_download_report()`:
```{r, eval=T, include=F, message=F, warning=F}
library(wildRtrax)

Sys.setenv(WT_USERNAME = 'guest', WT_PASSWORD = 'Apple123')
wt_auth()
data <- wt_download_report(605,'ARU',c('birdnet','main'))

```
```{r, eval=F, include=T, message=F, warning=F}
library(wildRtrax)

Sys.setenv(WT_USERNAME = 'guest', WT_PASSWORD = 'Apple123')
wt_auth()
data <- wt_download_report(605,'ARU',c('birdnet','main'))

```

The results from the `main` report can then be anti-joined against the `birdnet` report to determine if any species were detected than a human did not find. It's recommended to use high threshold values from BirdNET and species that should occur or be allowed in the project.

```{r}
library(magrittr)

birdnet_report <- data[[1]] %>%
  dplyr::filter(is_species_allowed_in_project == 't',
         confidence > 50) %>%
  tibble::add_column(observer = 'BirdNET') %>%
  dplyr::select(location, recording_date_time, observer, species_code) %>%
  dplyr::distinct()

main_report <- data[[2]] %>%
  wt_tidy_species(remove = c("mammal", "amphibian", "abiotic", "insect"), zerofill = F) %>%
  dplyr::select(location, recording_date_time, observer, species_code) %>%
  dplyr::distinct()

merged_data <- dplyr::bind_rows(birdnet_report, main_report) %>% dplyr::arrange(location, recording_date_time) %>%
  dplyr::group_by(location, recording_date_time, observer, species_code) %>%
  dplyr::mutate(presence = dplyr::row_number()) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = species_code, values_from = presence, values_fill = 0) %>%
  tidyr::pivot_longer(cols = -c(location, recording_date_time, observer), names_to = "species_code", values_to = "presence")

merged_data %>%
  dplyr::filter(observer == "BirdNET" & presence == 1) %>%
  dplyr::select(location, recording_date_time, species_code) %>%
  dplyr::anti_join(
    merged_data %>%
      dplyr::filter(observer != "BirdNET") %>%
      dplyr::select(location, recording_date_time, species_code, presence),
    by = c("location", "recording_date_time", "species_code")
  )

```

