---
title: "working with wildRtrax"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{working-with-wildRtrax}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(wildRtrax)
```


This document contains an outline for how to use the R package **wildRtrax** for environmental sensor data management and analytics. 

## **FAQ**

### What is **wildRtrax**?

**wildRtrax**, pronounced *'wilder tracks'*, is an R package for ecologists and advanced users who work with environmental sensor data, mainly from acoustic recordings units (ARUs) or remote game cameras (RGCs). It contains functions designed to meet most needs in order to organize, analyze and standardize data. **wildRtrax** is self-contained and must be run under an R statistical environment, and it also depends on many other R packages. **wildRtrax** is free software and distributed under [MIT License (c) 2020](https://github.com/mabecker89/wildRtrax/blob/master/LICENSE). All package functions are designed to standardize data to the [WildTrax](https://wwww.wildtrax.ca) infrastructure.

### What is **WildTrax**?

[**WildTrax**](https://www.wildtrax.ca) is a web-enabled portal to centralize, store, manage and share environmental sensor data. It was developed by the [Alberta Biodiversity Monitoring Institute](https://abmi.ca) and the [Bioacoustic Unit](https://bioacoustic.abmi.ca). **wildRtrax** serves as a parallel design and indicator to analytics and functions that can be developed in **WildTrax**. 

### What R packages does **wildRtrax** depend on?

**wildRtrax** depends on a multitude of packages to provide flexible routines and work flows for data management. [`tidyverse`](https://tidyverse.org) for piping functions, standard grammar and data manipulation, [`furrr`](https://davisvaughan.github.io/furrr/) and [`doParallel`](https://cran.r-project.org/web/packages/doParallel/index.html) for parallel computing, and acoustic analysis packages: [`bioacoustics`](https://cran.r-project.org/web/packages/bioacoustics/index.html), [`tuneR`](https://cran.r-project.org/web/packages/tuneR/index.html), [`seewave`](https://cran.r-project.org/web/packages/seewave/seewave.pdf). Certain functions depend on the [QUT Ecoacoustics Analysis Software](https://github.com/QutEcoacoustics/audio-analysis)

### Do I need to know any other programming langugages in order to use **wildRtrax**?

Certain functions rely on Python, Shell (bash) scripts and SQL. You do not need to know the languages in order to use the functions but it is strongly encouraged to read the documentation.

### How do I report a bug in **wildRtrax**?

If you think you have found a bug in **wildRtrax**, you should report it to developers or maintainers. Please do not send bug reports to R mailing lists, since **wildRtrax** is not a standard R package. The preferred forum to report bugs is [GitHub](https://github.com/mabecker89/wildRtrax/issues).  Here is what is required in order to report a bug - issues are welcome and encouraged and are the only way to make **wildRtrax** non-buggy:

* The bug report should be so detailed that the bug can be replicated and corrected
* Send an example that causes a bug
* Send a minimal dataset as well if it is not available in R
* Paste the output or error message in your message
* Specify which version of **wildRtrax** you used

### Can I contribute to **wildRtrax**?

Yes! **wildRtrax** is dependent on user contribution and all feedback is welcome. If you have problems with **wildRtrax**, it may be as simple as incomplete documentation. Feature requests also are welcome, but they are not necessarily fulfilled. A new feature will be added if it is easy to do and it looks useful to the user base of the package, or if you submit fully annotated code.
